---
title: "The impact of storms in the US"
author: "John Lee"
date: "24 September 2015"
output: html_document
---

Synopsis here: 10 sentences summarising analysis


# Data Processing
The following library's are used during the analaysis:
```{r}
library(data.table)
library(reshape2)
library(readr)
library(ggplot2)
```

The data used in this report is published by the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. The database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The data was downloaded from the following URL:
```{r, cache = TRUE}
file_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists("storm.csv.bz2")){
    download.file(file_url, destfile = "storm.csv.bz2",method = "curl")
    
}
```

The data was read using the read_csv() function from the readr package and converted to to the data.table datatype:
```{r, results = 'hide' }
storm_data <- data.table(read_csv("storm.csv.bz2"))
storm_data[,BGN_DATE:= as.POSIXct(BGN_DATE,format = "%m/%e/%Y %H:%M:%S")]
```

#Results
The data set has more than 900,000 entries with 37 variables for each entry:
```{r}
dim(storm_data)
```

```{r}
# p <- ggplot(storm_data[FATALITIES>20,], aes(EVTYPE, FATALITIES ))
# p + geom_boxplot() +
#     ggtitle("Fatalities") + 
#     ylab("Fatalities")

health_impact <- storm_data[,.(
    Total_fatalities = sum(FATALITIES),
    Total_injuries = sum(INJURIES)),
    by = EVTYPE][Total_fatalities>10]
injury_sum <- storm_data[,.(), by = EVTYPE][Total_injuries>10]

```

Hello
1-3 figures
    Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Across the United States, which types of events have the greatest economic consequences?
